---
title: "experiment"
author: "Maisy Hallam"
date: "2025-01-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lme4)
library(broom)
library(emmeans)

```


# Import data

```{r}
# how often participants generalised predictively
acc <- read_csv('data/experiment/ppt_accuracy.csv')

# simulated baseline MI for each participant
baseline <- read_csv('data/experiment/by_participant_baselines.csv')

# true MI for each participant
ppts <- read_csv('data/experiment/participant_mi.csv')
```


# Reshape data

Note: `system` numbers the kinship system types (Type I, IIgen, IIIfb etc) from 0 to 7.

0 = Type I
1 = Type IIside
2 = Type IIgen
3 = Type IIopp
4 = Type IIImb
5 = Type IIImz
6 = Type IIIfb
7 = Type IIIfz

```{r}
acc <- acc %>%
  mutate(
    # binary variable for whether a participant generalised predictively
    correct_01 = factor(correct),
    # factorise the system type
    system_type = factor(system_type)
    # and the system number
    system = factor(system)
    # factorise whether or not the ppt saw gender distinct given G0 terms
    gender_distinction_input = factor(gender_distinction),
    # group the systems by the kind of distinction made
    system_group = case_when(
      system == 0 ~ 'None',
      system %in% c(2,3) ~ 'Gender',
      system %in% c(1,4,5,6,7) ~ 'Other')
  )
```

# Models

Intercept-only mixed effects logistic regression predicting predictive generalisations.

```{r}
acc_mdl <- glmer(correct_01 ~ 1 + (1|random_id) + (1|system), family = 'binomial', data = acc)

summary(acc_mdl)
```

Mixed effects logistic regression predicting predictive generalisations by type of distinction (None, Gender, Other)

```{r}
acc_syst_mdl <- glmer(correct_01 ~ system_group + (1|random_id), family = 'binomial', data = acc)

summary(acc_syst_mdl)

emmeans(acc_syst_mdl, list(pairwise~"system_group"),adjust = 'bonferroni')
```

Mixed effects logistic regression predicting predictive generalisations by whether there
were gender distinctions in the G0 input.

```{r}
acc_gender_mdl <- glmer(correct_01 ~ gender_distinction_input + (1|random_id) + (1|system), family = 'binomial', data = acc)

summary(acc_gender_mdl)
```

# Mutual information analysis

Reshape the data to get z scores and how many participants produced maximum MI.

```{r}
all_bind <- rbind(baseline,ppts)

all <- merge(baseline,ppts, by = c('random_id','system')) %>%
  rename(simulated_value = mutual_information.x,
         true_value = mutual_information.y) 


ppt_z <- all %>% group_by(random_id,system) %>%
  summarise(
    mean = mean(simulated_value),
    sd = sd(simulated_value),
    true_value = true_value,
    z = (true_value - mean) / sd,
    significant = ifelse(z > 1.96 | z < -1.96, 'Y','N'),
    sim_min = round(min(simulated_value),5),
    sim_max = round(max(simulated_value),5),
    max = ifelse(round(true_value,2) == round(sim_max,2), 'Y','N'))
  ) %>%
  unique() # so we have one line per participant, not per baseline

```


Plot participants compared to their baselines.

```{r}
all %>% 
  merge(ppt_z, .by = c(random_id, system)) %>%
  mutate(z = ifelse(z == Inf, 0, z) %>% 
  na.omit() %>%
    ggplot(aes(x = true_value, y = random_id)) +
    geom_density_ridges(aes(x = simulated_value, y = random_id),alpha = 0.8, colour = '#666666',scale = 0.5) +
  scale_colour_manual(name = 'Significant?', values=c(Y = '#D0453B', N = 'black')) +
  geom_point(aes(colour = significant)) +
  facet_wrap(~system, nrow = 2, labeller = as_labeller(systems)) +
  labs(x = "Mutual information",
       y = "Participants") +
  theme(strip.text = ggtext::element_markdown(),
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_line(colour = 'grey'),
        axis.text.y = element_blank()
)
```


Percentage of systems produced by participants with significantly high mutual information.

```{r}
sum(z$z > 1.96) / count(z)
```

Percentage of systems produced by participants with significantly low mutual information. 
```{r}
sum(z$z < -1.96) / count(z)
```

Percentage of participants who produced maximum MI.

```{r}
sum(z$z > 1.96 && max == 'Y') / count(z)
```


# All plots

Proportion of predictive generalisations made by participants.

```{r}
all_accuracy_plt <- acc %>%
  
  group_by(random_id,system) %>%
  
  summarise(
    prop_correct = sum(correct) / n(),
    n_trials = n(),
    system = factor(system),
    system_type = system_type) %>%
  
  unique() %>%
  
  group_by() %>%
  
  mutate(
    mean_propn_expected = mean(prop_correct),
    sd = sd(prop_correct),
    n = n(),
    se = sd/sqrt(n),
    lower95 = mean_propn_expected - 1.96 * se,
    upper95 = mean_propn_expected + 1.96 * se
  ) %>%
  
  ggplot(aes(x = 1, y = prop_correct)) +
  
  geom_dotplot(fill = '#D0453B',binaxis = 'y', stackratio = 0.5, stackdir = 'center',    dotsize = 0.9, colour = '#D0453B',stroke = 0.01,alpha = 0.3) +
  
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.05, linewidth = 0.01,colour = 'black') +
  
  geom_point(aes(y = mean_propn_expected),size = 0.6) +
  labs(
       x = 'All participants',
       y = 'Proportion of \n predictive generalisations'
       ) +
  
  ylim(c(0,1)) +
  
  theme(legend.position = 'none',
        axis.text.x = element_blank(),
        text = element_text(size = 10))

all_accuracy_plt
```

Proportion of predictive generalisations grouped by distinction type.

```{r}
grouped_acc <- acc %>%  
  
  group_by(random_id, system, system_group) %>%
  
  summarise(
    prop_correct = sum(correct) / n(),
    n_trials = n()) %>%
  
  group_by(system_group) %>%
  
  summarise(
    prop_correct = prop_correct,
    mean_propn_expected = mean(prop_correct),
    sd = sd(prop_correct),
    n = n(),
    se = sd/sqrt(n),
    lower95 = mean_propn_expected - 1.96 * se,
    upper95 = mean_propn_expected + 1.96 * se,
    system_group = factor(system_group, levels = c('None','Non-Gender','Gender'))
  ) %>%
  
  ggplot(aes(x = system_group, y = prop_correct)) +
  
  scale_fill_manual(name = 'Distinctions amongst G+1 kin', labels = c('None','Non-Gender','Gender'),values=c('#D0453B','#E96922','#66748F')) +
  scale_colour_manual(name = 'Distinctions amongst G+1 kin', labels = c('None','Non-Gender','Gender'),values= c('#D0453B','#E96922','#66748F')) +

  geom_dotplot(aes(fill = system_group,colour = system_group),binaxis = 'y', stackratio = 0.4, stackdir = 'center', dotsize = 0.9,stroke = 0.01, alpha = 0.3) +
  
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.1, linewidth = 0.01,colour = 'black') +

  geom_point(aes(y = mean_propn_expected),size = 0.6) +
  labs(
       x = 'Distinctions amongst G+1 kin',
       y = ''
       ) +
  
  ylim(c(0,1)) +
  
  theme(legend.position = 'none',
        axis.text.x = element_text(size = 8),
        text = element_text(size = 10))

grouped_acc
```


Histogram of participants' z-scores.

```{r}
systems <- c('0' = 'Type I',
             '1' = 'Type II<sub>side</sub>',
             '2' = 'Type II<sub>gen</sub>',
             '3' = 'Type II<sub>opp</sub>',
             '4' = 'Type III<sub>fb</sub>',
             '5' = 'Type III<sub>mb</sub>',
             '6' = 'Type III<sub>fz</sub>',
             '7' = 'Type III<sub>mz</sub>')


ppt_z %>% group_by(system) %>% 
  mutate(
    insignificant = ifelse(z < 1.96 & z > -1.96,z,NA), 
    significant = ifelse(z > 1.96 | z < -1.96,z,NA)) %>% 
  
  ggplot() +
  
  geom_histogram(aes(x = insignificant), fill = 'grey',colour = '#666',size = 0.3,bins = 30) +
  geom_histogram(aes(x = significant), fill = '#ddafa9', colour = '#d95f50', size = 0.3,bins = 30) +
  
  labs(x = 'Mutual information (z-transformed)') +
  
  geom_vline(xintercept = 0,linetype = 'dashed',size = 0.2) +
  
  scale_x_continuous(breaks=c(-2,0,2,4),labels = c(-2,0,2,4)) +
  
  facet_wrap(~system, labeller = as_labeller(systems),nrow = 2) +
  
  theme(
    legend.position = 'none',
    strip.text = ggtext::element_markdown(),
    axis.text.x = element_text(size = 8))
```

Heatmaps depicting which referents were categorised together.

```{r}
ppt <- read_csv('data/experiment/participant_distance.csv')
ideal <- read_csv('data/experiment/ideal_distance.csv')

combined <- rbind(ppt,ideal)
```

Make facet names. `ppt` has system types listed as 1-8. `ideal` has system types listed as 9-16.

```{r}
facet_names <- c(
  '1' = 'Type~I',
'2' = 'Type~II[side]',
'3' = 'Type~II[gen]',
'4' = 'Type~II[opp]',
'5' = 'Type~III[fb]',
'6' = 'Type~III[fz]',
'7' = 'Type~III[mb]',
'8' = 'Type~III[mz]',
'9' = 'Type~I',
'10' = 'Type~II[side]',
'11' = 'Type~II[gen]',
'12' = 'Type~II[opp]',
'13' = 'Type~III[fb]',
'14' = 'Type~III[fz]',
'15' = 'Type~III[mb]',
'16' = 'Type~III[mz]'
)
```

```{r}
# colours for axis labels
colours = c('navy','darkred','navy','darkred','navy','darkred','navy','darkred','navy','darkred')

heatmaps <- combined %>%
  ggplot(aes(x = x, y = y, fill = z)) +
  geom_tile() +
  scale_fill_gradient(low="white", high="black") +
  labs(x = '', y = '') +
  theme(axis.text.x = element_text(colour = colours,angle = 90, vjust = 0.5, hjust = 1),
        axis.text.y = element_text(colour = colours),
        axis.text = element_text(size = 7),
        legend.position = 'none') +
  facet_wrap(~system,nrow = 2, labeller = as_labeller(facet_names, label_parsed))


heatmaps
```

